---
title: "Kfz-Neuzulassungen – Österreich"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Kfz-Neuzulassungen – Österreich}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # network calls – set TRUE locally to run
)
```

```{r setup}
library(davR)
library(dplyr)
```

## Overview

Three functions cover different aspects of Austrian Kfz-Neuzulassungen data.
All of them discover URLs dynamically — no manual link updates are needed when
Statistik Austria publishes new files each month.

| Function | Data | Coverage | Source |
|---|---|---|---|
| `statistik_get_kraftstoff_timeseries()` | PKW by fuel type × Bundesland | **2006-01 → today** | bundled Excel + live ODS |
| `statistik_get_kfz_neuzulassungen()` | All vehicle types by fuel type × Bundesland **or** brand × vehicle type | current year (monthly) or 2024–2025 (annual) | Statistik Austria ODS |
| `statistik_get_pkw_neuzulassungen()` | PKW by brand | **2000-01 → today** | STATcube OGD API (CSV) |

---

## 1 · `statistik_get_kraftstoff_timeseries()` — long time series by fuel type

The main workhorse for trend analysis. Returns monthly PKW registrations per
Bundesland and fuel/drive type from **January 2006** through the most recently
published month of the current year.

- **2006–2025**: served from the bundled `at_pkw_kraftstoff_hist` dataset
  (parsed from a Statistik Austria special evaluation).
- **Current year**: the current-year DE3 ODS file is discovered and downloaded
  fresh from the Statistik Austria website on every call.

### Columns

| column | type | description |
|---|---|---|
| `bundesland` | chr | Austrian state; `"Österreich"` = national total |
| `kraftstoffart` | chr | Fuel type (see below) |
| `year` | int | Year |
| `month` | int | Month (1–12) |
| `date` | Date | First day of the month |
| `anzahl` | int | New registrations (`NA` = suppressed, n < 5) |

### Available fuel types

```{r fuel_types}
at_pkw_kraftstoff_hist |>
  distinct(kraftstoffart) |>
  arrange(kraftstoffart)
```

### Full Elektro series

```{r ts_elektro}
statistik_get_kraftstoff_timeseries(kraftstoffart = "Elektro")
```

### Wien: Elektro vs. Diesel since 2020

```{r ts_wien}
statistik_get_kraftstoff_timeseries(
  date_from     = "2020-01",
  bundesland    = c("Wien", "Österreich"),
  kraftstoffart = c("Elektro", "Diesel")
)
```

### Historical data only (no network call)

```{r ts_hist_only}
statistik_get_kraftstoff_timeseries(include_current_year = FALSE)
```

---

## 2 · `statistik_get_kfz_neuzulassungen()` — richer current-year breakdown

Returns more detail for the current year (or annual summaries for 2024/2025):
per-vehicle-type breakdowns (PKW, Motorrad, LKW …) alongside Bundesland and
fuel type, **or** brand × vehicle type.

### `type = "kraftstoff"` — fuel × vehicle type × Bundesland

```{r kfz_kraft_cur}
# All months of the current year
statistik_get_kfz_neuzulassungen("kraftstoff")
```

```{r kfz_kraft_jan}
# Jänner only
df <- statistik_get_kfz_neuzulassungen("kraftstoff", months = "Jänner")

# PKW Elektro by Bundesland
df |>
  filter(fahrzeugtyp == "Personenkraftwagen Klasse M1",
         kraftstoffart == "Elektro") |>
  select(bundesland, anzahl) |>
  arrange(desc(anzahl))
```

```{r kfz_kraft_2025}
# Annual totals for 2025
statistik_get_kfz_neuzulassungen("kraftstoff", year = 2025)
```

### `type = "marken"` — brand × vehicle type (current year)

```{r kfz_marken}
dm <- statistik_get_kfz_neuzulassungen("marken")

# Top 10 PKW brands in Jänner
dm |>
  filter(monat == "Jänner",
         grepl("Personenkraftwagen", fahrzeugtyp),
         marke != "Kraftfahrzeuge insgesamt") |>
  slice_max(anzahl, n = 10) |>
  select(marke, anzahl)
```

### Available months and sheets

The current-year ODS contains one sheet per published month plus a cumulative
sheet (e.g. `"Jänner-März"`). Pass a character vector to `months` to select
specific sheets:

```{r kfz_months}
statistik_get_kfz_neuzulassungen(
  "kraftstoff",
  months = c("Jänner", "Februar", "März")
)
```

---

## 3 · `statistik_get_pkw_neuzulassungen()` — PKW by brand since 2000

Uses the **OGD REST API** (`data.statistik.gv.at`) to return monthly PKW brand
registrations from **January 2000** onward. Updated monthly.

### Columns

| column | type | description |
|---|---|---|
| `monat` | chr | Austrian label, e.g. `"Jänner 2026"` |
| `date` | Date | First day of the month |
| `marke` | chr | Brand name incl. country code, e.g. `"VW (D) <049400>"` |
| `anzahl` | int | New PKW registrations |

Note: Tesla appears as three separate rows (manufacturing in D, RC, USA). Sum
them for a combined Tesla figure.

### All brands, current year

```{r ogd_cur}
statistik_get_pkw_neuzulassungen(date_from = paste0(format(Sys.Date(), "%Y"), "-01"))
```

### Selected brands since 2015

```{r ogd_brands}
statistik_get_pkw_neuzulassungen(
  date_from = "2015-01",
  brands    = c("VW", "SKODA", "BMW", "MERCEDES", "TOYOTA", "TESLA")
)
```

### Top brands in the latest month

```{r ogd_top}
statistik_get_pkw_neuzulassungen() |>
  slice_max(date, n = 1, by = marke, with_ties = FALSE) |>
  slice_max(anzahl, n = 15) |>
  select(date, marke, anzahl)
```

---

## Combining the functions

A common workflow is to pair the long fuel-type series with the brand data:

```{r combine}
# Share of Elektro registrations by Bundesland over time
elektro <- statistik_get_kraftstoff_timeseries(
  date_from     = "2019-01",
  kraftstoffart = c("Elektro", "Benzin", "Diesel",
                    "Benzin/Elektro (hybrid)", "Diesel/Elektro (hybrid)")
)

elektro_share <- elektro |>
  summarise(anzahl = sum(anzahl, na.rm = TRUE),
            .by    = c(bundesland, kraftstoffart, date)) |>
  mutate(share = anzahl / sum(anzahl, na.rm = TRUE),
         .by   = c(bundesland, date)) |>
  filter(kraftstoffart == "Elektro")
```

---

## Data sources

| source | URL |
|---|---|
| Statistik Austria – Kfz-Neuzulassungen page | <https://www.statistik.at/statistiken/tourismus-und-verkehr/fahrzeuge/kfz-neuzulassungen> |
| OGD API – PKW by brand | <https://data.statistik.gv.at/web/meta.jsp?dataset=OGD_fkfzul0759_OD_PkwNZL_1> |
| Bundled historical Excel (special evaluation) | `system.file("extdata", "at_pkw_kraftstoff_hist_2006_2025.xlsx", package = "davR")` |

License: CC BY 4.0 – Statistik Austria.
